---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# b3gbi: B-Cubed General Biodiversity Indicators

Analyze biodiversity trends and spatial patterns from GBIF data cubes, using flexible indicators like richness, evenness, and more.

<!-- badges: start -->

[![Development Status:: In Development](https://img.shields.io/badge/Status-In%20Development-yellow.svg)](https://github.com/your_github_username/b3gbi)

<!-- badges: end -->

## Motivation

Biodiversity researchers need robust and standardized tools to analyze the vast amounts of data available on platforms like GBIF. The b3gbi package leverages the power of data cubes to streamline biodiversity assessments. It helps researchers gain insights into:

* **Changes Over Time:**  How biodiversity metrics shift throughout the years.
* **Spatial Variations:** Differences in biodiversity across regions, identifying hotspots or areas of concern. 
* **The Impact of Factors:** How different environmental variables or human activities might affect biodiversity patterns. 

## Key Features

b3gbi empowers biodiversity analysis with:

* **Standardized Workflows:**  Simplify the process of calculating common biodiversity indicators from GBIF data cubes.
* **Flexibility:**  Calculate richness, evenness, rarity, taxonomic distinctness, Shannon-Hill diversity, Simpson-Hill diversity, and more.
* **Analysis Options:** Explore temporal trends or create spatial maps.
* **Visualization Tools:** Generate publication-ready plots of your biodiversity metrics.

## Installation

You can install the development version of b3gbi from [GitHub](https://github.com/) with:

``` {r}
# install.packages("devtools")
# devtools::install_github("shawndove/b3gbi")
```

## Example

This is a basic example which shows you how to calculate and plot a map of species richness for a data cube containing GBIF occurrence data on amphibians in Europe:

```{r example}
# Load necessary libraries
library(b3gbi)
library(data.table)
library(ggplot2)

# Load GBIF data cube
cube_name <- "inst/extdata/europe_amphibians_cube.csv"

# Load taxonomic info for cube
tax_info <- "inst/extdata/europe_amphibians_info.csv"

# Prepare cube
amphib_data <- process_cube(cube_name, tax_info)

# Calculate diversity metric
map_obs_rich_amphib <- calculate_indicator(amphib_data, type = "obs_rich")

# Plot diversity metric
plot(map_obs_rich_amphib, title = "Observed Species Richness: Amphibians in Europe") 
```


You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
