---
title: "Uncertainty in Biodiversity Indicators"
author: "Shawn Dove"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncertainty in Biodiversity Indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "95%"
)
```

## Introduction 🛡️

Quantifying uncertainty is crucial for robust biodiversity assessments. The **b3gbi** package provides a flexible and powerful way to calculate confidence intervals (CIs) for biodiversity indicators using bootstrapping methods, leveraging the **dubicube** package for robust cube-level resampling.

To maintain a clean and efficient workflow, uncertainty calculation is decoupled from the initial indicator calculation. This "two-step" process allows users to first explore their data and indicators quickly, and then perform computationally intensive bootstrapping only when needed.

## The Two-Step Workflow

The standard workflow for adding uncertainty to an indicator is as follows:

1.  **Calculate the Indicator**: Use a time-series wrapper function (e.g., `total_occ_ts()`, `pielou_evenness_ts()`).
2.  **Add Confidence Intervals**: Pass the resulting indicator object to the `add_ci()` function.

### Example: Calculating Richness with Uncertainty

In this example, we calculate total occurrences for mammals in Denmark and then add 95% confidence intervals using cube-level bootstrapping.

```{r example, eval = FALSE}
library(b3gbi)

# 1. Process the cube
denmark_cube <- process_cube(system.file("extdata", 
                                         "denmark_mammals_cube_eqdgc.csv", 
                                         package = "b3gbi"))

# 2. Calculate a time series of total occurrences
# By default, CIs are NOT calculated during this step
occ_ts <- total_occ_ts(denmark_cube)

# 3. Add confidence intervals using add_ci()
# This step uses the dubicube package for robust bootstrapping
occ_ts_with_ci <- add_ci(occ_ts, num_bootstrap = 100) # Using 100 for speed in this example

# 4. Plot the result
plot(occ_ts_with_ci, title = "Total Occurrences with 95% CI")
```

## Bootstrapping Levels

The `add_ci()` function supports two levels of bootstrapping, selectable via the `bootstrap_level` argument:

### 1. Cube-Level Bootstrapping (`bootstrap_level = "cube"`)

This is the **default and recommended method**. It resamples the raw occurrence records within the data cube. 

*   **Pros**: Mathematically more robust; captures the underlying sampling uncertainty of the original data.
*   **Cons**: Computationally more intensive, especially for large cubes.

It leverages the `dubicube` package to ensure that indicators are recalculated correctly for each bootstrap replicate.

### 2. Indicator-Level Bootstrapping (`bootstrap_level = "indicator"`)

This method resamples the calculated indicator values themselves.

*   **Pros**: Very fast, even for massive datasets.
*   **Cons**: Less robust; assumes that the calculated indicator values are independent and identically distributed, which is often not strictly true for biodiversity metrics.

## Advanced Configuration

The `add_ci()` function provides several arguments for fine-tuning the CI calculation:

| Argument | Description | Default |
|----------|-------------|---------|
| `num_bootstrap` | Number of bootstrap replicates. | `1000` |
| `ci_type` | Type of bootstrap interval (`"norm"`, `"basic"`, `"perc"`, `"bca"`). | `"norm"` |
| `confidence_level` | The confidence level (e.g., 0.95). | `0.95` |
| `boot_args` | A list of additional arguments for `dubicube::bootstrap_cube()`. | `list()` |
| `ci_args` | A list of additional arguments for `dubicube::calculate_bootstrap_ci()`. | `list()` |

### Customizing the Bootstrap Process

If you need to pass specific parameters to the underlying `dubicube` functions (e.g., setting a specific seed), you can use the `boot_args` and `ci_args` parameters:

```{r advanced, eval = FALSE}
occ_ts_custom <- add_ci(occ_ts, 
                        num_bootstrap = 500,
                        boot_args = list(seed = 42),
                        ci_args = list(type = "perc"))
```

## Supported Indicators

Confidence intervals can be added post-hoc for the following indicators:

*   Total Occurrences (`total_occ`)
*   Occurrence Density (`occ_density`)
*   Newness (`newness`)
*   Williams Evenness (`williams_evenness`)
*   Pielou Evenness (`pielou_evenness`)
*   Abundance-based Rarity (`ab_rarity`)
*   Area-based Rarity (`area_rarity`)
*   Species Occurrences (`spec_occ`)
*   Species Range (`spec_range`)

### Exclusions

Certain indicators cannot have confidence intervals added via `add_ci()`:

*   **Hill Numbers** (`hill0`, `hill1`, `hill2`): These calculate their own uncertainty internally using the `iNext` package during the initial calculation.
*   **Observed Richness** (`obs_richness`): Bootstrapping observed richness is often not statistically sensible; consider using Hill numbers for estimated richness instead.
*   **Cumulative Richness** (`cum_richness`): The cumulative nature of this metric makes standard bootstrapping inappropriate.
*   **Occurrence Turnover** (`occ_turnover`): Similar to cumulative richness, the temporal dependency requires specialized methods.
*   **Taxonomic Distinctness** (`tax_distinct`): Requires specialized randomization tests rather than standard bootstrapping.

## Data Exploration with dubicube 🔍

While **b3gbi** focuses on indicator calculation and visualization, the underlying **dubicube** package offers powerful functions for initial data exploration and quality assessment of your biodiversity data cubes.

To learn more about the full capabilities of **dubicube**, including in-depth tutorials and advanced data processing techniques, please visit the [dubicube documentation](https://docs.b-cubed.eu/software/dubicube/readme/).

## Summary

The `add_ci()` function provides a unified and robust interface for adding uncertainty to your biodiversity assessments. By leveraging the power of `dubicube`, **b3gbi** ensures that your results are not just numbers, but scientifically grounded estimates with clearly defined confidence boundaries.
