---
title: "Uncertainty in Biodiversity Indicators"
author: "Shawn Dove"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncertainty in Biodiversity Indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "95%"
)
```

## Introduction 🛡️

Quantifying uncertainty is crucial for robust biodiversity assessments. The **b3gbi** package provides a flexible and powerful way to calculate confidence intervals (CIs) for biodiversity indicators using bootstrapping methods, leveraging the **dubicube** package for robust cube-level resampling.

To maintain a clean and efficient workflow, uncertainty calculation is decoupled from the initial indicator calculation. This "two-step" process allows users to first explore their data and indicators quickly, and then perform computationally intensive bootstrapping only when needed.

## The Two-Step Workflow

The standard workflow for adding uncertainty to an indicator is as follows:

1.  **Calculate the Indicator**: Use a time-series wrapper function (e.g., `total_occ_ts()`, `pielou_evenness_ts()`).
2.  **Add Confidence Intervals**: Pass the resulting indicator object to the `add_ci()` function.

### Example: Calculating Richness with Uncertainty

In this example, we calculate total occurrences for mammals in Denmark and then add 95% confidence intervals using cube-level bootstrapping.

```{r example, eval = FALSE}
library(b3gbi)

# 1. Process the cube
denmark_cube <- process_cube(system.file("extdata", 
                                         "denmark_mammals_cube_eqdgc.csv", 
                                         package = "b3gbi"))

# 2. Calculate a time series of total occurrences
# By default, CIs are NOT calculated during this step
occ_ts <- total_occ_ts(denmark_cube)

# 3. Add confidence intervals using add_ci()
# This step uses the dubicube package for robust bootstrapping
occ_ts_with_ci <- add_ci(occ_ts, num_bootstrap = 100) # Using 100 for speed in this example

# 4. Plot the result
plot(occ_ts_with_ci, title = "Total Occurrences with 95% CI")
```

## Bootstrapping Levels

The `add_ci()` function supports two levels of bootstrapping, selectable via the `bootstrap_level` argument:

### 1. Cube-Level Bootstrapping (`bootstrap_level = "cube"`)

This is the **default and recommended method**. It resamples the raw occurrence records within the data cube. The function automatically determines whether to use group-specific resampling (for species-level indicators) or whole-cube resampling (for aggregate indicators) based on the indicator type.

*   **Pros**: Mathematically more robust; captures the underlying sampling uncertainty of the original data.
*   **Cons**: Computationally more intensive, especially for large cubes.

It leverages the `dubicube` package to ensure that indicators are recalculated correctly for each bootstrap replicate.

### 2. Indicator-Level Bootstrapping (`bootstrap_level = "indicator"`)

This method resamples the calculated indicator values themselves.

*   **Pros**: Very fast, even for massive datasets.
*   **Cons**: Less robust; assumes that the calculated indicator values are independent and identically distributed, which is often not strictly true for biodiversity metrics.

## Automatic Indicator Configuration

The `add_ci()` function automatically applies appropriate bootstrapping strategies based on the indicator type. This ensures statistically correct confidence intervals without requiring manual configuration:

*   **Group-Specific Bootstrapping**: Species-level indicators (`total_occ`, `spec_occ`, `spec_range`) automatically resample within each species-year combination. This is essential when indicators are calculated per species.

*   **Whole-Cube Bootstrapping**: Aggregate indicators (evenness, rarity, density metrics) resample across the entire cube, treating the data as a single population.

*   **Bounded Indicators**: Evenness metrics (`pielou_evenness`, `williams_evenness`) automatically use logit transformation to ensure confidence intervals remain within valid [0, 1] bounds.

These defaults are applied internally, but can be overridden using `boot_args` if needed.

## Advanced Configuration

The `add_ci()` function provides several arguments for fine-tuning the CI calculation:

| Argument | Description | Default |
|----------|-------------|---------|
| `num_bootstrap` | Number of bootstrap replicates. | `1000` |
| `ci_type` | Type of bootstrap interval (`"norm"`, `"basic"`, `"perc"`, `"bca"`). | `"norm"` |
| `confidence_level` | The confidence level (e.g., 0.95). | `0.95` |
| `boot_args` | A list of additional arguments for `dubicube::bootstrap_cube()`. | `list()` |
| `ci_args` | A list of additional arguments for `dubicube::calculate_bootstrap_ci()`. | `list()` |

### Customizing the Bootstrap Process

If you need to pass specific parameters to the underlying `dubicube` functions (e.g., setting a specific seed), you can use the `boot_args` and `ci_args` parameters:

```{r advanced, eval = FALSE}
occ_ts_custom <- add_ci(occ_ts, 
                        num_bootstrap = 500,
                        boot_args = list(seed = 42),
                        ci_args = list(type = "perc"))
```

### Overriding Indicator Defaults

While `add_ci()` automatically selects appropriate settings, you can override these defaults using `boot_args` and `ci_args`. This is useful when you need specialized behavior:

```{r override, eval = FALSE}
# Example: Calculate CIs for evenness on raw scale (no logit transformation)
evenness_raw <- add_ci(my_evenness_indicator,
                       num_bootstrap = 1000,
                       boot_args = list(trans = identity, 
                                        inv_trans = identity))

# Example: Force bias correction for total_occ
occ_with_bias <- add_ci(my_occ_indicator,
                        num_bootstrap = 1000,
                        ci_args = list(no_bias = FALSE))
```

Common override parameters include:
- `trans` / `inv_trans`: Transformation functions
- `no_bias`: Logical, disable bias correction (default varies by indicator)
- `group_specific`: Logical, force group-specific vs whole-cube bootstrapping

## Supported Indicators

Confidence intervals can be added post-hoc for the following indicators:

*   Total Occurrences (`total_occ`)
*   Occurrence Density (`occ_density`)
*   Newness (`newness`)
*   Williams Evenness (`williams_evenness`)
*   Pielou Evenness (`pielou_evenness`)
*   Abundance-based Rarity (`ab_rarity`)
*   Area-based Rarity (`area_rarity`)
*   Species Occurrences (`spec_occ`)
*   Species Range (`spec_range`)

### Exclusions

Certain indicators cannot have confidence intervals added via `add_ci()`:

*   **Hill Numbers** (`hill0`, `hill1`, `hill2`): These calculate their own uncertainty internally using the `iNext` package during the initial calculation.
*   **Observed Richness** (`obs_richness`): Bootstrapping observed richness is often not statistically sensible; consider using Hill numbers for estimated richness instead.
*   **Cumulative Richness** (`cum_richness`): The cumulative nature of this metric makes standard bootstrapping inappropriate.
*   **Occurrence Turnover** (`occ_turnover`): Similar to cumulative richness, the temporal dependency requires specialized methods.
*   **Taxonomic Distinctness** (`tax_distinct`): Requires specialized randomization tests rather than standard bootstrapping.

## Data Exploration with dubicube 🔍

While **b3gbi** focuses on indicator calculation and visualization, the underlying **dubicube** package offers powerful functions for initial data exploration and quality assessment of your biodiversity data cubes.

To learn more about the full capabilities of **dubicube**, including in-depth tutorials and advanced data processing techniques, please visit the [dubicube documentation](https://docs.b-cubed.eu/software/dubicube/readme/).

## Summary

The `add_ci()` function provides a unified and robust interface for adding uncertainty to your biodiversity assessments. By leveraging the power of `dubicube`, **b3gbi** ensures that your results are not just numbers, but scientifically grounded estimates with clearly defined confidence boundaries.

With automatic indicator configuration, most users can rely on sensible defaults while advanced users retain full control over the bootstrapping process through the `boot_args` and `ci_args` parameters.
